---
title: "visualizeGO - Tutorial"
author: "Alonso-García, M."
date: "2025-06-23"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducción a visualizeGO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```
# Introduction
The ```visualizeGO``` package provides enhanced visualizations for Gene Ontology (GO) enrichment analysis results. This vignette shows how to use ```visualizeGO``` with ```clusterProfiler``` to run a GO enrichment analysis and visualize the results.

# Load packages
```{r libraries}
library(visualizeGO)
library(clusterProfiler)
```
# Examples
## Case 1: Role of alveolar luminal cells in the mouse mammary gland during pregnancy

## Case 2: Functional genomics of the digestive tract in broilers.

## Case 3: Hypomethylation in bull sperm targets specific genomic functions


```{r pressure, echo=FALSE}
plot(pressure)
```
